Final Project

### 2025-05-03 MSSC5750_Project.qmd
Update MSSC5750_Project.qmd

>>>>

## ADD CODE FOR LDA

library(tidyverse)
library(MASS)
bankr<-data.frame(loan_data$person_age,loan_data$person_income,loan_data$person_emp_exp,loan_data$loan_amnt,
                  loan_data$loan_int_rate,loan_data$loan_percent_income,loan_data$cb_person_cred_hist_length,
                  loan_data$credit_score,loan_data$previous_loan_defaults_on_file)

head(bankr)

lda_desc<-lda(loan_data$previous_loan_defaults_on_file~loan_data$person_age+
                loan_data$person_income + loan_data$person_emp_exp+loan_data$loan_amnt +
                loan_data$loan_int_rate+ loan_data$loan_percent_income+loan_data$cb_person_cred_hist_length+
                loan_data$credit_score,data=bankr )

lda_desc

a<-lda_desc$scaling
a

library(MVN)
library(tidyverse)

previous_default<-bankr%>%
  filter(loan_data.previous_loan_defaults_on_file=="Yes")%>%
  select(-loan_data.previous_loan_defaults_on_file)
  s1<-cov(previous_default)

  s1

previous_default
# Checking for multivariate normality
defaulterscheck<-mvn(previous_default[,1:8],mvnTest = "mardia")$multivariateNormality
print(defaulterscheck)

#separate groups into two and compute sample covariance matrix
no_default<-bankr%>%
  filter(loan_data.previous_loan_defaults_on_file =="No")%>%
  select(-loan_data.previous_loan_defaults_on_file)
s2<-cov(no_default)
s2

# pooled sample covariance matrix  is :

n1<-nrow(previous_default)

n2<-nrow(no_default)

Spl<-1/(n1+n2-2)*((n1-1)*s1 +(n2-1)*s2)

Spl

# manually calculating the coefficients of the standardized discriminant function

a.star<-sqrt(diag(Spl))*a

a.star
# a.star ranks the independependent variables in the order of discriminative capacity
# i.e which variables best separate the groups: load_default on file vs no history of default

# 1 age
#2. income
#3. employment experience
#4. loan amount
#5. loan interest rate
# 6. loan percent of income (i.e debt-to-income ratio)
#7.  length of credit history
# 8. Credit score

library(tidyverse)
ggplot(bankr,aes(x=loan_data$person_age,y=loan_data$loan_percent_income,
                 color=factor(loan_data$previous_loan_defaults_on_file)))+
  geom_point()+ labs(xlab="Age",ylab = "debt-to-incomes")+
  scale_color_discrete(name="loan_data$previous_loan_defaults_on_file")

non_defaultcheck<-mvn(no_default[,1:8],mvnTest = "mardia")$multivariateNormality

print(non_defaultcheck)

pca_bank<-prcomp(bankr[,1:8],scale. = TRUE)
summary(pca_bank)

pca_bank # pca to confirm the variability in data set and ranking of variables 
           # in discrimination  b/w 2 groups

lda2<-lda(loan_data.previous_loan_defaults_on_file~loan_data.person_age + loan_data.person_income+
            loan_data.person_emp_exp +loan_data.loan_amnt,data =bankr)

lda2

library(MASS)
library(tidyverse)
library(Hotelling)

loans.test<-hotelling.test(loan_data$person_age+
    loan_data$person_income + loan_data$person_emp_exp+loan_data$loan_amnt +
    loan_data$loan_int_rate+ loan_data$loan_percent_income+loan_data$cb_person_cred_hist_length+
    loan_data$credit_score~factor(loan_data$previous_loan_defaults_on_file),data=bankr )

loans.test

#LDA-Predict groups-taking a predictive view

lda3<-lda(loan_data$previous_loan_defaults_on_file~loan_data$person_age+
            loan_data$person_income + loan_data$person_emp_exp+loan_data$loan_amnt +
            loan_data$loan_int_rate+ loan_data$loan_percent_income+loan_data$cb_person_cred_hist_length+
            loan_data$credit_score,prior=c(0.5,0.5),data=bankr )
lda3.results<-bankr%>%
  mutate(classify.loan_data.previous_loan_defaults_on_file=predict(lda3)$class)

lda3.results

lda3$scaling

zbar1 <- sum(lda3$scaling*lda3$means[1,])
zbar1

zbar2<- sum(lda3$scaling*lda3$means[2,])

zbar2

cutoff<-0.5*(zbar1+zbar2)

cutoff

#checking classification results
with(lda3.results,table(loan_data.previous_loan_defaults_on_file,classify.loan_data.previous_loan_defaults_on_file))

#Classification Analysis Result

# overall classification  (13,508 + 14793)/45,000 : 63% Correct classification

# No default :13508/(13,508+8,634)

# Previous default : 14,793/(14,793+8,065)

pairs(bankr[,1:8])
